### 2018-02-09 submission/manuscript.Rmd
update abstract

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness, after correcting for study effect and variable region sequenced, with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in the odds ratio for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples. A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples. Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

**Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma genesis. The evidence collected within this study on the role of the microbiota in CRC identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.

@@

microbiota; colorectal cancer; polyps; adenoma; meta-analysis.

>>>>

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased, after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in the odds ratio for individuals to have both adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma tumors (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for individuals to have carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased odds ratios for individuals to have either adenoma or carcinoma tumors regardless of sample type. Prediction models built to classify either individuals with adenoma or carcinoma were trained on the whole community or selected genera (top 5 highest and lowest odds ratios) from either fecal or tissue samples. Both the full and select models for either adenoma or carcinoma resulted in similar classification success according to Area Under the Curve (AUC). Importantly, all models had a generalized decrease in AUC between train and test data sets. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and tumor were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

**Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma genesis. The evidence collected within this study on the role of the microbiota in those with tumors identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.

@@

microbiota; colorectal cancer; polyps; adenoma; tumor; meta-analysis.

### 2018-02-09 submission/manuscript.tex
update abstract

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness, after correcting for study effect and
variable region sequenced, with more severe disease (P-value \textless{}
0.05). This reduction in evenness translated into small increases in the
odds ratio for adenoma (P-value = 0.035) and carcinoma stages of CRC
(P-value = 0.0021) while the reduction in Shannon diversity only
translated into an increased odds ratio for developing carcinomas
(P-value = 0.0069). Increases in mouth-associated microbes were commonly
in the top 5 most significantly increased relative risk of adenoma and
carcinoma for both stool and tissue samples. A prediction model for
adenoma and carcinoma was built using either the whole community or
selected genera with highest and lowest relative risk from fecal and
tissue samples. Both approaches resulted in similar classification
success according to Area Under the Curve (AUC) regardless of whether
genera or OTUs were used to build the model. The most important groups
within the full community models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Although a number of associations between the microbiota and
CRC were identified, the majority of studies that we used in this
meta-analysis were only individually adequately powered for large effect
sizes.

\textbf{Conclusions.} These data provide support for the importance of
the bacterial community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
CRC identifies a number of correlations that may not have been detected
because of the low power associated with the majority of studies that
have been performed to date.

>>>>

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness as tumor severity increased, after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness translated into small
increases in the odds ratio for individuals to have both adenoma
(P-value = 0.035) and carcinoma tumors (P-value = 0.0021) while the
reduction in Shannon diversity only translated into an increased odds
ratio for individuals to have carcinomas (P-value = 0.0069). Increases
in mouth-associated microbes were commonly in the top 5 most
significantly increased odds ratios for individuals to have either
adenoma or carcinoma tumors regardless of sample type. Prediction models
built to classify either individuals with adenoma or carcinoma were
trained on the whole community or selected genera (top 5 highest and
lowest odds ratios) from either fecal or tissue samples. Both the full
and select models for either adenoma or carcinoma resulted in similar
classification success according to Area Under the Curve (AUC).
Importantly, all models had a generalized decrease in AUC between train
and test data sets. The most important groups within the full community
models consistently belonged to genera such as \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} across studies. Although a
number of associations between the microbiota and tumor were identified,
the majority of studies that we used in this meta-analysis were only
individually adequately powered for large effect sizes.

\textbf{Conclusions.} These data provide support for the importance of
the bacterial community to both adenoma and carcinoma genesis. The
evidence collected within this study on the role of the microbiota in
those with tumors identifies a number of correlations that may not have
been detected because of the low power associated with the majority of
studies that have been performed to date.
